\section{Basic Data Types}
There are two kinds of data types in C\#.

\begin{itemize}
    \item Value Types (implicit data types, structs and enumeration)
    \item Reference Types (objects, delegates)  
\end{itemize}

Value types are passed to methods by passing an exact copy while Reference types are passed to methods by
passing only their reference (handle). Implicit data types are defined in the language core by the language vendor,
while explicit data types are types that are made by using or composing implicit data types.
The commonly used implicit data types in C\# are mentioned below:

\begin{center}
    \begin{tabular}{ | m{5em} | m{2cm}| m{10cm} | } 
    \hline
    Data Type & Size & Description \\
    \hline
    int	    &   4 bytes &	Stores whole numbers from -2,147,483,648 to 2,147,483,647\\
    long	&	8 bytes	&   Stores whole numbers from -9,223,372,036,854,775,808 to 9,223,372,036,854,775,807\\
    float	&	4 bytes	&   Stores fractional numbers. Sufficient for storing 6 to 7 decimal digits\\
    double	&	8 bytes	&   Stores fractional numbers. Sufficient for storing 15 decimal digits\\
    decimal	&	16 bytes&   Stores fractional numbers. Sufficient for storing 28 decimal digits\\
    bool	&	1 byte	&   Stores true or false values\\
    char	&	2 bytes	&   Stores a single character/letter, surrounded by single quotes\\
    string	&	2 bytes per character &	Stores a sequence of characters, surrounded by double quotes\\
    \hline
    \end{tabular}
\end{center}


Implicit data types are represented in language using keywords, so each of the above is a keyword in C\#. Keyword
are the words defined by the language and can not be used as identifiers. It is worth noting that string is also an
implicit data type in C\#, so string is a keyword in C\#. The last point about implicit data types is that they are value
types and thus stored on the stack, while user defined types or referenced types are stored using the heap. A stack is
a data structure that store items in a first in first out (FIFO) fashion. It is an area of memory supported by the
processor and its size is determined at the compile time. A heap consists of memory available to the program at run
time. Reference types are allocated using memory available from the heap dynamically (during the execution of
program). The garbage collector searches for non-referenced data in heap during the execution of program and
returns that space to Operating System.\\