\section{Variables}
During the execution of a program, data is temporarily stored in memory. A variable is the name given to a
memory location holding a particular type of data. So, each variable has associated with it a data type and a value.

In C\#, variables are declared as:\\

\colorbox{lightgray}{<data type> <variable>;}\\

e.g.,\\
\colorbox{lightgray}{int i;}\\


The above line will reserve an area of 4 bytes in memory to store an integer type values, which will be referred to
in the rest of program by the identifier ’i’. You can initialize the variable as you declare it (on the fly) and can also
declare/initialize multiple variables of the same type in a single statement, e.g.,

\colorbox{lightgray}{bool isReady = true;}\\
\colorbox{lightgray}{float percentage = 87.88, average = 43.9;}\\
\colorbox{lightgray}{char digit = '7';}\\

In C\# (like other modern languages), you must declare variables before using them. Also, there is the concept of
"Definite Assignment" in C\# which says "local variables (variables defined in a method) must be initialized before
being used". The following program won’t compile:

\begin{lstlisting}
    static void Main()
    {
        int age;
        // age = 18;
        Console.WriteLine(age); // error
    }
    
\end{lstlisting}

But, if you un-comment the 2nd line, the program will compile. C\# does not assign default values to local
variables. C\# is also a type safe language, i.e., values of particular data type can only be stored in their respective
(or compatible) data type. You can’t store integer values in Boolean data types like we used to do in C/C++.


\textbf{Constant Variables or Symbols}

Constants are variables whose values, once defined, can not be changed by the program. Constant variables are
declared using the const keyword, like:\\

\colorbox{lightgray}{const double PI = 3.142;}\\


Constant variables must be initialized as they are declared. It is a syntax error to write:

\colorbox{lightgray}{const int MARKS;}\\

It is conventional to use capital letters when naming constant variables.

\textbf{Naming Conventions for variables and methods}

Microsoft suggests using Camel case Notation (first letter in lowercase) for variables and Pascal Notation (first letter in
uppercase) for methods. Each word after the first word in the name of both variables and methods should start with
a capital letter. For example, variable names following Camel notation could be:

salary totalSalary
myMathsMarks isPaid

Some typical names of method following Pascal Notation are

GetTotal() Start()
WriteLine() LastIndexOf()

Although it is not mandatory to follow this convention, it is highly recommended that you strictly follow the
convention.